Mixing time estimation in reversible Markov chains from a single sample path

### 2015-06-02 abstract.tex
Mostly abstract/intro
Making abstract more precise.
New reference.
\ln -> \log replacement (should be the other way around?)
Changed some section titles

%!TEX root =  matrix-est.tex
This article provides an efficient procedure for estimating the mixing
time of a reversible ergodic Markov chain.
The procedure computes the estimate from a single sample path of the
Markov chain, and does not require any \emph{a priori} bounds on the
mixing time itself.
This stands in contrast to previous approaches, which either require a
reset mechanism or require additional prior knowledge.
Most notably, the procedure produces a fully empirical confidence
interval (i.e., it depends only on the sample path and a given
confidence parameter $\delta$) that is always valid with probability
$1-\delta$; and the width of the interval converges to zero at a
$\sqrt{n}$-rate, where $n$ is the length of the sample path.

>>>>

%!TEX root =  matrix-est.tex
This article provides the first efficient, non-heuristic procedure for computing a fully data dependent interval
that traps the mixing time of a finite reversible ergodic Markov chain with a prescribed probability.
The interval is computed from a single finite-length sample
path from the Markov chain, and does not require the 
knowledge of any parameters of the chain.
%bounds on the mixing time itself.
This stands in contrast to previous approaches, which either 
only provide point estimates, or require a
reset mechanism, or additional prior knowledge.
%Most notably, 
%the procedure produces a fully empirical confidence
%interval (i.e., it depends only on the sample path and a given
%confidence parameter $\delta$) that is always valid with probability
%$1-\delta$; and 
The width of the interval converges to zero at a
$\sqrt{n}$ rate, where $n$ is the length of the sample path.
Upper and lower bounds are given on the number of samples required
to achieve a constant accuracy estimate in terms of the chain's parameters,
in particular, based on the absolute spectral gap $\gap$.
\todoc{We may need to add some explanation to connect $\gap$ estimates
to $\tmix$ estimates.}
Our lower bounds indicate that, 
unless further restrictions are placed on the chain,
no procedure can achieve a constant
accuracy before seeing each state at least  $\Omega(1/\gap)$ times.
Future directions of research are identified.
\todoc{Such as: closing the gap. Extension to large state spaces, 
based on additional assumptions. Even if $\gap$ is large, we need many samples.
Maybe we need even stronger assumptions, like the decay rate of eigenvalues.
}
%If $\gap$ denotes the absolute spectral gap of the unknown Markov chain
%and the chain's stationary distribution is minorized by $\pimin$,
%our procedure requires $\tilde{O}( \frac{1}{\pimin\gap^3})$ observations
% to estimate the mixing time up to a constant multiplicative accuracy, 
%while we also establish that any algorithm to achieve the same
%goal would need $\Omega( d\log(d)/\gap + 1/\pimin)$ samples.

%and the dependence on the main  chain's

### 2015-06-02 intro-nips.tex
Mostly abstract/intro
Making abstract more precise.
New reference.
\ln -> \log replacement (should be the other way around?)
Changed some section titles

This work considers the construction of fully empirical bounds on the
mixing time of Markov chains based on a single sample path.
Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic
time-homogeneous Markov chain on a finite state space $[d] :=
\{1,2,\dotsc,d\}$ with transition matrix $\vP$.
Under this assumption, the chain converges to its unique stationary
\todoc{Maybe steady state is better?} distribution $\vpi =
(\pi_i)_{i=1}^d$ regardless of the initial state distribution $\vq$: 
\[
  \lim_{t\to\infty} {\Pr}_{\vq}\Parens{X_t = i}
  = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i
  \quad \text{for each $i \in [d]$} .
\]
The mixing time $\tmix$ of the Markov chain is the time that it takes
for the chain to (approximately) reach its stationary distribution,
\[
  \tmix
  :=
  \min\Braces{
    t \in \bbN :
    \sup_{\vq}
    \max_{A\subset [d]}
    \Abs{
      \textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)
    }
%    \Norm{
%      \vq \vP^t - \vpi
%    }_{\tv} 
    \leq 1/4
  }\,.
\]
Here, $\vpi(A) = \sum_{i\in A} \pi_i$ is the probability assigned to
set $A$ by $\vpi$, and the supremum is over all possible initial
distributions $\vq$.
The problem studied in this work is the construction of a non-trivial
confidence interval $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset
[0,\infty]$, based only on the observed sample path
$(X_1,X_2,\dotsc,X_n)$ and $\delta \in (0,1)$, that captures the
mixing time $\tmix$ with the pre-specified probability $1-\delta$.

@@

\paragraph{Main results.}
We develop the first efficient procedure for constructing non-trivial
and fully empirical confidence intervals for Markov mixing time.
Consider a reversible ergodic Markov chain on $d$ states with absolute
spectral gap $\gap$ and stationary distribution minorized by $\pimin$.
(These quantities $\gap$ and $\pimin$ are well-known to govern
mixing time.)
Our main results are summarized as follows.
\begin{enumerate}
  \item
    In \cref{thm:err}, we give an estimator for $\gap$ that
    achieves multiplicative accuracy from \emph{a single sample path}
    of length $\tilde{O}\parens{1/(\pimin\gap^3)}$.\footnote{The
    $\tilde{O}(\cdot)$ notation suppresses logarithmic factors.}
    We also provide an estimator for $\pimin$ that requires a sample
    path of length $\tilde{O}\parens{1/(\pimin\gap)}$.
    This establishes the feasibility of \emph{estimating} the mixing
    time in this setting.
    However, the valid confidence intervals suggested by
    \cref{thm:err} depend on the unknown quantities $\pimin$ and
    $\gap$.

  \item
    In \cref{thm:lb-pimin,thm:lb-gap}, we show that $\Omega\parens{
    (d\log d)/\gap + 1/\pimin}$ observations are necessary for any
    procedure to guarantee constant multiplicative accuracy in
    estimating $\gap$.
    Essentially, every state may need to be visited about
    $\log(d)/\gap$ times, on average, before an accurate estimate of
    the mixing time can be provided.

  \item
    In \cref{thm:empirical}, we develop a procedure for computing
    \emph{valid empirical confidence intervals} for $\pimin$ and
    $\gap$ (\cref{alg:empest}).
    The procedure uses just a single sample path and does not require
    any prior knowledge about $\pimin$ or $\gap$.
    We also quantify the rate at which the width of the confidence
    intervals shrink as a function of the length of the sample path.

  \item
    Finally, in \cref{thm:combined}, we show how to combine the
    empirical confidence intervals from \cref{alg:empest} with the
    non-empirical bounds from \cref{thm:err} to produce valid
    empirical confidence intervals;
    the width of these intervals converge to zero asymptotically as
    fast as those from \cref{thm:err}.

%    which is
%    $     O\Parens{
%      \frac1{\pimin}\sqrt{\frac{d\ln(d/\delta)}{ n}}
%      +
%      \frac{d}{\pimin\gap} \sqrt{\frac{\ln(d/\delta)}{\pimin n}}
%    }.$

@@

Although these asymptotic results help us understand the limiting
behavior of the sample mean over a Markov chain, they say little about
the finite-time non-asymptotic behavior, which is often needed for the
prudent evaluation of a method or even its algorithmic design
\cite{
MCMCDiscussion93%
,DBLP:conf/valuetools/KontoyiannisLM06%
,BBL06%
,MniSzeAu08%
,MauPo09%
,LiLiWaSt11:KWIK%
,flegal2011implementing%
,Gyori-paulin15%
,SwaJoa15:LoggedBandit%
}. %BaHaVa10:activelearning,ShaVo08:conformalpred
To address this need, numerous works have developed Chernoff-type
bounds on $\Pr\parens{ |\vpi_n(f)-\vpi(f)| > \eps }$, thus providing
invaluable tools for non-asymptotic probabilistic
analysis~\cite{gillman1998chernoff,leon2004optimal,DBLP:conf/valuetools/KontoyiannisLM06,paulin15}.
These probability bounds are larger than corresponding bounds for
independent and identically distributed (iid) data due to the temporal
dependence; intuitively, for the Markov chain to yield a fresh draw
$X_{t'}$ that behaves as if it was independent of $X_t$, one must wait
$\Theta(\tmix)$ time steps.
Note that the bounds generally depend on distribution-specific
properties of the Markov chain (e.g., $\vP$, $\tmix$), which are often
unknown \emph{a priori} in practice.
Consequently, much effort has been put towards estimating these
unknown quantities, especially in the context of MCMC diagnostics, in
order to provide data-dependent assessments of estimation
accuracy~\cite[e.g.,][]{MCMCDiscussion93,jones2001,flegal2011implementing,1209.0703,Gyori-paulin15}.
However, these approaches generally only provide asymptotic
guarantees, and hence fall short of our goal of empirical bounds that
are valid with any finite-length sample path.

Many results from statistical learning and empirical process theory
have been extended to sufficiently fast mixing
data~\citep[e.g.,][]{Yu94,MR1921877,gamarnik03,MoRo08,MoRo09,DBLP:conf/nips/SteinwartC09,Steinwart2009175},
providing learnability assurances (e.g., generalization error bounds)
in scenarios where data are dependent.
These results are often given in terms of mixing coefficients, which
can be consistently estimated in some cases~\citet{McDoShaSche11}.
However, the convergence rates of the estimates
from~\citet{McDoShaSche11}, which are needed to derive confidence
bounds, are given in terms of unknown mixing coefficients.
When the data comes from a Markov chain, these mixing coefficients can
often be bounded in terms of mixing times, and hence our main results
provide a way to make them fully empirical.

It is possible to eliminate many of the difficulties presented above
with additional access to the Markov chain.
For example, with access to a sampling oracle that generates
independent transitions from any given state (akin to a ``reset''
device), the mixing time becomes an efficiently testable property in
the sense studied in~\citet{BaFoRuSmiWhi00,BaFoRuSmiWhi13}.
On the other hand, when one only has a circuit-based description of
the transition probabilities of Markov chain over an
exponentially-large state space, there are complexity-theoretic
barriers for many MCMC diagnostic problems~\citet{BhaBoMo11}.

@@

%    (d\ln d)/\gap + 1/\pimin }$ observations are necessary for

@@

%%      \frac1{\pimin}\sqrt{\frac{d\ln(d/\delta)}{ n}}

@@

%%      \frac{d}{\pimin\gap} \sqrt{\frac{\ln(d/\delta)}{\pimin n}}

@@

%% $\exp(W(\ln(n)))$ is the window length, where $W$ is Lambert's $W$-function (i.e., this is between $\ln \ln(n)$ and $\ln n$).

>>>>

This work considers the construction of fully empirical bounds on the
mixing time of Markov chains based on a single sample path.
Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic
time-homogeneous Markov chain on a finite state space $[d] :=
\{1,2,\dotsc,d\}$ with transition matrix $\vP$.
Under this assumption, the chain converges to its unique stationary
\todoc{Maybe steady state is better?} distribution $\vpi =
(\pi_i)_{i=1}^d$ regardless of the initial state distribution $\vq$: 
\[
  \lim_{t\to\infty} {\Pr}_{\vq}\Parens{X_t = i}
  = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i
  \quad \text{for each $i \in [d]$} .
\]
The mixing time $\tmix$ of the Markov chain is the time that it takes
for the chain to (approximately) reach its stationary distribution,
\begin{align}
\label{eq:mixtimedef}
  \tmix
  :=
  \min\Braces{
    t \in \bbN :
    \sup_{\vq}
    \max_{A\subset [d]}
    \Abs{
      \textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)
    }
%    \Norm{
%      \vq \vP^t - \vpi
%    }_{\tv} 
    \leq 1/4
  }\,.
\end{align}
Here, $\vpi(A) = \sum_{i\in A} \pi_i$ is the probability assigned to
set $A$ by $\vpi$, and the supremum is over all possible initial
distributions $\vq$.
The problem studied in this work is the construction of a non-trivial
confidence interval $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset
[0,\infty]$, based only on the observed sample path
$(X_1,X_2,\dotsc,X_n)$ and $\delta \in (0,1)$, that captures the
mixing time $\tmix$ with the pre-specified probability $1-\delta$.

@@

\paragraph{Main results.}
We develop the first efficient \todoc{Maybe leave out efficient? 
What is efficient is a question of interpretation. We 
should maybe rather add a note on the computational complexity.}
procedure for constructing non-trivial
and fully empirical confidence intervals for Markov mixing time.
Consider a reversible ergodic Markov chain on $d$ states with absolute
spectral gap $\gap$ and stationary distribution minorized by $\pimin$. \todoc{Comment on reversibility somewhere?}
As is well known \citep[Theorems~12.3 and~12.4]{LePeWi08},
$(\frac{1}\gap  - 1)\log(2)  \le \tmix \le \frac{1}\gap \log \frac4\pimin$,
hence it suffices to estimate $\gap$ and $\pimin$. 
\todoc{I would prefer $\ln$ over $\log$ throughout the paper.}
%these quantities $\gap$ and $\pimin$ are well-known to govern
%mixing time.
Our main results are summarized as follows.
\begin{enumerate}
  \item
    In \cref{thm:err}, we give a point-estimator for $\gap$ that
    achieves multiplicative accuracy from \emph{a single sample path}
    of length $\tilde{O}\parens{1/(\pimin\gap^3)}$.\footnote{The
    $\tilde{O}(\cdot)$ notation suppresses logarithmic factors.}
    We also provide a point-estimator for $\pimin$ that requires a sample
    path of length $\tilde{O}\parens{1/(\pimin\gap)}$.
    This establishes the feasibility of \emph{estimating} the mixing
    time in this setting.
    However, the valid confidence intervals suggested by
    \cref{thm:err} depend on the unknown quantities $\pimin$ and
    $\gap$.

  \item
    In \cref{thm:lb-pimin,thm:lb-gap}, we show that $\Omega\parens{
    (d\log d)/\gap + 1/\pimin}$ observations are necessary for any
    procedure to guarantee constant multiplicative accuracy in
    estimating $\gap$.
    Essentially, \emph{every} state may need to be visited about
    $\log(d)/\gap$ times, on average, before an accurate estimate of
    the mixing time can be provided, regardless of what
    estimation procedure one uses.

  \item
    In \cref{thm:empirical}, we develop a procedure for computing
    \emph{valid empirical confidence intervals} for $\pimin$ and
    $\gap$ (\cref{alg:empest}).
    The procedure uses just a single sample path and does not require
    any prior knowledge of the chain's parameters, such as $\pimin$ or $\gap$.
    We also quantify the rate at which the width of the confidence
    intervals shrink as a function of the length of the sample path.

  \item
    Finally, in \cref{thm:combined}, we show how to combine the
    empirical confidence intervals from \cref{alg:empest} with the
    non-empirical bounds from \cref{thm:err} to produce valid
    empirical confidence intervals;
    the width of these intervals converge to zero asymptotically as
    fast as those from \cref{thm:err}.
    \todoc{Is it true from the construction of \cref{thm:combined}
    that if we had a better point estimator
    then we would automatically get an interval estimator
    with the same accuracy?
    This seems to be true.
    This would mean that to close the gap
    one should really work on coming up with better point estimators.
    }

%    which is
%    $     O\Parens{
%      \frac1{\pimin}\sqrt{\frac{d\log(d/\delta)}{ n}}
%      +
%      \frac{d}{\pimin\gap} \sqrt{\frac{\log(d/\delta)}{\pimin n}}
%    }.$

@@

Although these asymptotic results help us understand the limiting
behavior of the sample mean over a Markov chain, they say little about
the finite-time non-asymptotic behavior, which is often needed for the
prudent evaluation of a method or even its algorithmic design
\cite{
MCMCDiscussion93%
,DBLP:conf/valuetools/KontoyiannisLM06%
,BBL06%
,MniSzeAu08%
,MauPo09%
,LiLiWaSt11:KWIK%
,flegal2011implementing%
,Gyori-paulin15%
,SwaJoa15:LoggedBandit%
}. %BaHaVa10:activelearning,ShaVo08:conformalpred
To address this need, numerous works have developed Chernoff-type
bounds on $\Pr\parens{ |\vpi_n(f)-\vpi(f)| > \eps }$, thus providing
invaluable tools for non-asymptotic probabilistic
analysis~\cite{gillman1998chernoff,leon2004optimal,DBLP:conf/valuetools/KontoyiannisLM06,paulin15}.
These probability bounds are larger than corresponding bounds for
independent and identically distributed (iid) data due to the temporal
dependence; intuitively, for the Markov chain to yield a fresh draw
$X_{t'}$ that behaves as if it was independent of $X_t$, one must wait
$\Theta(\tmix)$ time steps.
Note that the bounds generally depend on distribution-specific
properties of the Markov chain (e.g., $\vP$, $\tmix$, $\gap$), which are often
unknown \emph{a priori} in practice.
Consequently, much effort has been put towards estimating these
unknown quantities, especially in the context of MCMC diagnostics, in
order to provide data-dependent assessments of estimation
accuracy~\cite[e.g.,][]{MCMCDiscussion93,GaSmi00:eigval,jones2001,flegal2011implementing,1209.0703,Gyori-paulin15}.
However, these approaches generally only provide asymptotic
guarantees, and hence fall short of our goal of empirical bounds that
are valid with any finite-length sample path.

Learning with dependent data is another main motivation to our work.
Many results from statistical learning and empirical process theory
have been extended to sufficiently fast mixing, dependent
data~\citep[e.g.,][]{Yu94,MR1921877,gamarnik03,MoRo08,MoRo09,DBLP:conf/nips/SteinwartC09,Steinwart2009175},
providing learnability assurances (e.g., generalization error bounds).
These results are often given in terms of mixing coefficients, which
can be consistently estimated in some cases~\citet{McDoShaSche11}.
However, the convergence rates of the estimates
from~\citet{McDoShaSche11}, which are needed to derive confidence
bounds, are given in terms of unknown mixing coefficients.
When the data comes from a Markov chain, these mixing coefficients can
often be bounded in terms of mixing times, and hence our main results
provide a way to make them fully empirical, at least in the limited setting we study.

It is possible to eliminate many of the difficulties presented above
with additional access to the Markov chain.
For example, with access to a sampling oracle that generates
independent transitions from any given state (akin to a ``reset''
device), the mixing time becomes an efficiently testable property in
the sense studied in~\citet{BaFoRuSmiWhi00,BaFoRuSmiWhi13}.
On the other hand, when one only has a circuit-based description of
the transition probabilities of Markov chain over an
exponentially-large state space, there are complexity-theoretic
barriers for many MCMC diagnostic problems~\citet{BhaBoMo11}.
\todoc{Explain the organization of the paper?}

@@

%    (d\log d)/\gap + 1/\pimin }$ observations are necessary for

@@

%%      \frac1{\pimin}\sqrt{\frac{d\log(d/\delta)}{ n}}

@@

%%      \frac{d}{\pimin\gap} \sqrt{\frac{\log(d/\delta)}{\pimin n}}

@@

%% $\exp(W(\log(n)))$ is the window length, where $W$ is Lambert's $W$-function (i.e., this is between $\log \log(n)$ and $\log n$).

### 2015-06-02 notation.tex
adding and resolving todos, as per our discussion.
bib file improvements
generalized inverse clean up, both in appendix and in main text
ideas for future work in conclusion
empirical results: comment on O(.) notation
minor tweak to intro
Mostly abstract/intro
Making abstract more precise.
New reference.
\ln -> \log replacement (should be the other way around?)
Changed some section titles

%!TEX root =  matrix-est.tex
We denote the set of positive integers by $\bbN$, and
the set of the first $d$ positive integers $\{1,2,\dotsc,d\}$ by $\iset{d}$.
The non-negative part of a real number $x$ is $[x]_+ := \max\{0,x\}$.
Boldface symbols are used for vectors and matrices (e.g., $\vv$,
$\vM$), and their entries are referenced by subindexing (e.g., $v_i$,
$M_{i,j}$).
%The $i$-th coordinate basis vector is denoted by $\ve_i$.
For a vector $\vv$, $\norm{\vv}$ denotes its Euclidean norm; for a
matrix $\vM$, $\norm{\vM}$ denotes its spectral norm.
We use $\Diag(\vv)$ to denote the diagonal matrix whose $(i,i)$-th
entry is $v_i$.
The probability simplex is denoted by $\Delta^{d-1} = \{ \vp
\in [0,1]^d : \sum_{i=1}^d p_i = 1 \}$, and we regard vectors in
$\Delta^{d-1}$ as row vectors.
Finally, the total variation between two probability distributions
$\vp, \vq \in \Delta^{d-1}$ is denoted by $\norm{\vp-\vq}_{\tv}$.

>>>>

%!TEX root =  matrix-est.tex
We denote the set of positive integers by $\bbN$, and
the set of the first $d$ positive integers $\{1,2,\dotsc,d\}$ by $\iset{d}$.
The non-negative part of a real number $x$ is $[x]_+ := \max\{0,x\}$,
$\log$ denotes the natural logarithm.
Boldface symbols are used for vectors and matrices (e.g., $\vv$,
$\vM$), and their entries are referenced by subindexing (e.g., $v_i$,
$M_{i,j}$).
%The $i$-th coordinate basis vector is denoted by $\ve_i$.
For a vector $\vv$, $\norm{\vv}$ denotes its Euclidean norm; for a
matrix $\vM$, $\norm{\vM}$ denotes its spectral norm.
We use $\Diag(\vv)$ to denote the diagonal matrix whose $(i,i)$-th
entry is $v_i$.
The probability simplex is denoted by $\Delta^{d-1} = \{ \vp
\in [0,1]^d : \sum_{i=1}^d p_i = 1 \}$, and we regard vectors in
$\Delta^{d-1}$ as row vectors.
%Finally, the total variation between two probability distributions
%$\vp, \vq \in \Delta^{d-1}$ is denoted by $\norm{\vp-\vq}_{\tv}$.
